Metabolites reflect the mesoscale variability in biogeochemistry introduced by sea-surface eddies in the North Pacific Subtropical Gyre

### 2023-09-20 manuscript/manuscript.Rmd
Finished discussion and polished up a few todo's
Ready to send to collaborators imo

---
title: "MESOSCOPE manuscript"
author: "William Kumler"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template.docx
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE)
library(tidyverse)

@@

@@

These recent works have highlighted the effect of eddies on both bulk oceanographic parameters such as particulate carbon and nitrogen and the composition of the microbial community, but we still lack direct links between the genomic and transcriptomic assays occurring regularly around the world and the particulate organic matter produced by those organisms [@Kharbush2020]. This missing link is bridged by metabolomics, a nascent field studying the structure and function of the cell on a molecular level. Of particular interest within the metabolome are those small, polar molecules that represent the fundamental unit of cell biology and function as building blocks for macromolecules [@Moran2022]. These small molecules also have roles of their own within the cell, acting as osmolytes, antioxidants, nutrient storage compounds, and more [@Geraldes2021; @Mojzes2020; @Kharbush2020]. Despite these critical functions, they are difficult to study in the marine environment due to their low concentrations and similarity to the abundant salt ions, resulting in challenges in detection and quantification [@Boysen2018; @Moran2022; @Sacks2022].

@@

  - Discussion: note that the "neither" samples all group together, implying eddy effect rather than location ("outside" the eddy is more similar to each other than to the insides of either one)
  - Discussion: it's kinda crazy that the exact center of the eddy is so wildly different at the DCM from stations that are nearby but not

@@

# Permutation testing for effect size test on Deep SLA differences
set.seed(123)
map_dbl(seq(1, 2, 0.1), function(n_mult){
  map_dbl(1:100, function(h){
    pc_comp_df %>%
      filter(depth=="Deep") %>%
      group_by(sla_class) %>%
      summarize(mean=mean(frac_in_metab), sd=sd(frac_in_metab), n=n()) %>%
      mutate(sim_vals=map(1:3, function(i){rnorm(.$n[i]*n_mult, .$mean[i], .$sd[i])})) %>%
      unnest_longer(sim_vals) %>%
      aov(formula=sim_vals~sla_class) %>%
      broom::tidy() %>%
      filter(term=="sla_class") %>%
      pull(p.value)
  }) %>% mean()
})
```

This correlation between PC and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. Known metabolites contributed between approximately 1.5% and 5% of the carbon biomass to the total PC load, but this variance was not clearly due to either depth or SLA (Supplemental figure 2). The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.3-0.4 percentage points lower at depth, p-value=0.018) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). However, this lack of difference observed may have been due to the small sample sizes (18 anticyclonic samples, 9 cyclonic, and 6 neither). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for the 175 meter samples if 1.5x as many samples had been collected with mean and variances remaining similar.

  - Discussion: why are my % PC in metabs lower than Angie's? Because her PC values were way higher but metab concentrations were about the same. Wait, are they even?

@@

@@

```{r regression relative abundance coefficients, eval=FALSE}
reg_coefs <- untarg_peaks %>%
  # filter(feature!="MS_NEG_0155") %>%
  left_join(filled_file_metadata %>% distinct(filename, ms_run, station, depth, sla_corr)) %>%
  filter(ms_run=="MT") %>%
  filter(depth%in%c("Surface", "DCM", "Deep")) %>%
  filter(feature%in%unique(feature)) %>%
  group_by(filename) %>%
  mutate(norm_area=norm_area/sum(norm_area)) %>%
  ungroup() %>%
  nest(data=-c(depth, feature)) %>%
  mutate(test=map(data, ~broom::tidy(lm(.x$norm_area~.x$sla_corr)), .progress = TRUE)) %>%
  unnest(test)
reg_coefs %>%
  filter(str_detect(term, "sla_corr")) %>%
  mutate(p.value=p.adjust(p.value, method="fdr")) %>%
  arrange(p.value) %>%
  left_join(untarg_feats) %>%
  select(feature, depth, estimate, p.value, mzmed, rtmed, compound_name) %>%
  filter(p.value<0.05) %>%
  # as.data.frame() %>%
  filter(depth=="Deep") %>% 
  print(n=Inf)
reg_coefs %>%
  filter(str_detect(term, "sla_corr")) %>%
  left_join(untarg_feats) %>%
  filter(compound_name=="Acetylcholine?") %>%
  # filter(feature=="MS_POS_1395") %>%
  filter(depth=="Deep") %>%
  select(compound_name, depth, data, estimate) %>%
  unnest(data) %>%
  group_by(sla_corr, station) %>%
  ggplot(aes(x=sla_corr, y=norm_area*100)) +
  geom_smooth(method="lm", formula="y~x") +
  geom_boxplot(aes(group=sla_corr)) +
  scale_y_continuous(limits = c(0, NA))
```

@@

Finally, in the 175 meter samples we detected 32 mass features with a significant association with SLA. Notably, all detected nucleobases (guanine, adenine, and cytosine) were positively associated with SLA (higher values in the anticyclone than the cyclone), though the strongest associations were found in O-acetylcarnitine, betonicine, and tyrosine. Both O-acetylcarnitine and betonicine effectively quintupled their contribution to total peak area, shifting from approximately 0.1% to ~0.5% over the transect. Acetylcholine was the only known compound more relatively abundant in the cyclone than the anticyclone along with one unknown of *m/z* 131.034. The unknown at *m/z* 173.09211 with the strongest trend at the DCM again had the strongest trend in the 175 meter samples, here with an FDR-corrected p-value of $2.0*10^{-7}$.

  - Discussion: surprised that so many compounds are still different after normalizing within each file - expected that to remove biomass signal
- Discussion: why are homarine and trigonelline so correlated?
- Discussion: homarine/trigonelline has a larger effect size but lower p-value because of nonlinear fit (exponential rise into eddy center rather than linear)
- Discussion: how can most compounds be relatively enriched in the anticyclone at 175 meters? Where is the remainder of the effect size going?
- Discussion: all the nucleobases upregulated in the 175m anticyclone, why? Bacteria?

@@

To provide context for the sea level anomaly responses detected above, we additionally performed high-resolution depth sampling around the DCM at the center of both eddy poles. Here, we found that the ~50 meter difference in depth and the ~40 centimeter difference in SLA explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001). Additionally, both sample depth and SLA were well-correlated with the first principal component of the metabolite matrix (r$_{depth}=.766$, r$_{SLA}=.684$, fraction of variance explained by PC$_1$ = 30.8%) (Figure 3).

```{r figure MC NMDS plot}
MC_nmds_gp <- multivar_analysis_MC$nmds_output %>%
  pluck(1, "points") %>% 
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_class, depth)) %>%
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"),
                      labels=c("-20m", "-10m", "DCM", "+10m", "+20m"))) %>%
  ggplot() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, color=sla_class), size=4, pch=19) +
  scale_color_manual(
    values=c(scales::muted("blue"), scales::muted("red")),
    breaks=c("Cyclone", "Anticyclone"),
  ) +
  annotate("label", x=-200, y=100, label="Cyclone", color=scales::muted("blue"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=100, y=120, label="Anticyclone", color=scales::muted("red"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=-Inf, y=Inf, label=multivar_analysis_MC$nmds_stress,
           hjust=0, vjust=1, label.r = unit(0, "in")) +
  geom_point(aes(x=MDS1, y=MDS2, fill=depth, alpha=depth), size=4, pch=21, color="black") +
  scale_fill_manual(
    values = c("grey90", "grey80", "grey60", "grey40", "grey20"),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.2, 0, 0.2, 0.4),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  coord_fixed() +
  guides(alpha="none", color="none", fill=guide_legend(
    title = "Depth", title.position = "top", title.hjust = 0.5, direction = "horizontal"
  )) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave("MC_nmds_gp.png", plot = MC_nmds_gp, device = "png", 
       path = "figures", width = 6.5, height = 4.5, units = "in", dpi = 300)
```

@@

  - Discussion: discuss how the cyclone+20m is pretty similar to the anticyclone samples and why
  - Discussion: less of a clear depth trend despite the stats saying equal?
  - Discussion: two possible models, one where samples from the same absolute depth are similar, one where samples from the same distance from DCM are similar. Can I argue for which one is better here?

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 4). We found that a majority of the metabolites (Clusters 1 and 2 in Figure 4, 63% of the total) fell into clusters with larger peak areas in the cyclone, as expected, while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 4). Cluster 2 notably had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster 1 also showed a neat DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (r=.851) that was not present in the cyclone (r=.013).

```{r k-means clustering}
set.seed(123)
kclust <- untarg_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MC") %>%
  select(feature, filename, norm_area) %>%
  group_by(feature) %>%
  filter(sum(norm_area==0)<n()*0.9) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  pivot_wider(names_from = "filename", values_from = norm_area) %>%
  column_to_rownames("feature") %>%
  data.matrix() %>%
  kmeans(centers = 4)

@@

```{r eddy center volcano plot}
volcanoplot_gp <- pairwise_tests %>%
  left_join(fold_change_df, by = join_by(feature)) %>%
  left_join(rownames_to_column(data.frame(cluster=kclust$cluster), "feature"), 
            by = join_by(feature)) %>%
  left_join(untarg_feats %>% select(feature, compound_name), by = join_by(feature)) %>%
  mutate(known=ifelse(is.na(compound_name), "Unknown", "Known")) %>%
  mutate(compound_name=ifelse(is.na(compound_name), feature, compound_name)) %>%
  ggplot(aes(x=log2(fold_change), y=-log10(p.value), color=factor(cluster), 
             shape=known, label=compound_name)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 2, lty=2) +
  geom_point(size=2, stroke=1) +
  scale_y_continuous(breaks = 0:6, labels = paste0("10<sup>-", 0:6, "</sup>")) +
  scale_shape_manual(values=c(4, 16), breaks = c("Unknown", "Known")) +
  scale_color_manual(
    values = c("#414487FF", "#7AD151FF", "#22A884FF", "#2A788EFF"),
    breaks = 1:4
  ) +
  theme_bw() +
  labs(x="Log-2 fold-change cyclone enrichment", y="Mann-Whitney FDR-corrected p-value",
       shape="Metabolite\nstatus", color="k-means\ncluster") +
  guides(color=guide_legend(order = 2), shape=guide_legend(order=1)) +
  theme(axis.text.y = ggtext::element_markdown())

ggsave("volcanoplot_gp.png", plot = volcanoplot_gp, device = "png", 
       path = "figures", width = 6.5, height = 3, units = "in", dpi = 300)
```

![](figures/volcanoplot_gp.png)

>>>>

---
title: "Metabolites reflect the mesoscale variability in biogeochemistry introduced by sea-surface eddies in the North Pacific Subtropical Gyre"
author: "William Kumler"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template.docx
bibliography:
- Exported Items.bib
- r-packages.bib
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)
library(RaMS)

@@

TBD

@@

These recent works have highlighted the effect of eddies on both bulk oceanographic parameters such as particulate carbon and nitrogen and the composition of the microbial community, but we still lack direct links between the genomic and transcriptomic assays occurring regularly around the world and the particulate organic matter produced by those organisms [@Kharbush2020]. This missing link is bridged by metabolomics, a nascent field studying the structure and function of the cell on a molecular level. Of particular interest within the metabolome are those small, polar molecules that represent the fundamental unit of cell biology and function as building blocks for macromolecules [@Moran2022]. These small molecules also have roles of their own within the cell, acting as osmolytes, antioxidants, nutrient storage compounds, and more [@Mojzes2020; @Kharbush2020]. Despite these critical functions, they are difficult to study in the marine environment due to their low concentrations and similarity to the abundant salt ions, resulting in challenges in detection and quantification [@Boysen2018; @Moran2022; @Sacks2022].

@@

@@

# Permutation testing for effect size test on Deep SLA differences
set.seed(123)
map_dbl(seq(1, 2, 0.1), function(n_mult){
  map_dbl(1:100, function(h){
    pc_comp_df %>%
      filter(depth=="Deep") %>%
      group_by(sla_class) %>%
      summarize(mean=mean(frac_in_metab), sd=sd(frac_in_metab), n=n()) %>%
      mutate(sim_vals=map(1:3, function(i){rnorm(.$n[i]*n_mult, .$mean[i], .$sd[i])})) %>%
      unnest_longer(sim_vals) %>%
      aov(formula=sim_vals~sla_class) %>%
      broom::tidy() %>%
      filter(term=="sla_class") %>%
      pull(p.value)
  }) %>% mean()
}, .progress = TRUE)
```

This correlation between PC and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. Known metabolites contributed between approximately 1.5% and 5% of the carbon biomass to the total PC load, but this variance was not clearly due to either depth or SLA (Supplemental figure 2). The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.6 percentage points lower at 175 meters, p-value=0.002) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). However, this lack of difference observed may have been due to the small sample sizes (18 anticyclonic samples, 9 cyclonic, and 6 neither). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for SLA in the 175 meter samples if ~1.5x as many samples had been collected with mean and variances remaining similar.

  - Discussion: My % PC in metabs lower than Angie's? Because her PC values were way higher but metab concentrations were about the same?

@@

### A single sample highlights a network of related metabolites

One sample, taken from the surface at Station 11 (MS11C215m_A), was highly distinct from all others (including its paired triplicates) with several metabolites measuring two or more orders of magnitude larger in both peak area and concentration. Heuristic exploration of nearby masses identified a total of 14 compounds highly abundant in this one sample alone. The only identified compounds with this trend were taurine and acetyltaurine, though three others with *m/z* 146.1176, 148.0968, and 206.1388 were discovered to be likely isomers of known compounds (acetylcholine/(3-carboxypropyl)trimethylammonium, hydroxyisoleucine, dexpanthenol, respectively). One feature (*m/z* = 184.0637, RT = 7.0 minutes) was identified as choline sulfate using a second run of standards, while an additional mass feature at 168.0689 and RT = 8.5 was putatively flagged as taurine betaine (same mass as hydroxyisoleucine). This mass was also one of the most highly enriched in the MS11C215m_A sample after taurine itself, with a peak area nearly 30x that of the next largest sample. No clear explanation emerged explaining which compounds were hyperabundant in this surface sample, though one possible explanation was the *Trichodesmium* bloom occurring at the same time.

```{r MS11A offenders table}
ms11_offenders <- untarg_peaks %>%
  filter(str_detect(filename, "180821")) %>%
  group_by(feature) %>%
  filter(sum(norm_area==0)<3) %>%
  mutate(cursed_area=norm_area[filename=="180821_Smp_MS11C215m_A.mzML"]) %>%
  filter(cursed_area>0) %>%
  mutate(rel_area=norm_area/cursed_area) %>%
  filter(filename!="180821_Smp_MS11C215m_A.mzML") %>%
  group_by(feature) %>%
  summarise(mean_rel=mean(rel_area), max_rel=max(rel_area)) %>%
  # ggplot(aes(x=mean_rel, y=max_rel, label=feature)) + geom_point() + 
  # scale_y_log10() + scale_x_log10() +
  # geom_vline(xintercept = 0.25) + geom_hline(yintercept = 0.5)
  filter(mean_rel<0.25 & max_rel < 0.5) %>%
  arrange(mean_rel) %>% 
  add_row(feature="MS_NEG_1123") %>%
  left_join(untarg_feats %>% select(feature:compound_name)) %>%
  mutate(mzmed=ifelse(str_detect(feature, "NEG"), mzmed+1.007276*2, mzmed)) %>%
  filter(feature!="MS_NEG_0768") %>% # Remove taurine duplicate
  add_row(mzmed=c(198.0794, 184.0637), rtmed=c(5, 7))

# ms11_offenders$mzmed %>% sort() %>% `names<-`(., round(., digits=3)) %>% dist(diag = TRUE) %>% round(digits=4)

stans <- read_csv("https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv") %>%
  filter(Column=="HILIC" & z>0) %>%
  select(cmpd_name=Compound_Name, formula=Empirical_Formula) %>%
  group_by(formula) %>%
  summarise(cmpd_name=paste(cmpd_name, collapse = ", "))
base_mz <- 85.089149355+1.007276
iter_mzs <- expand_grid(CH2=0:6, O=2:6) %>%
  mutate(mz=base_mz+CH2*14.01565+O*15.994915) %>%
  mutate(mz=round(mz, digits=5)) %>%
  mutate(cmd=paste0("mz%between%pmppm(", mz, ", 10)")) %>%
  mutate(formula=paste0("C", 5+CH2, "H", 11+CH2*2, "NO", O)) %>%
  left_join(stans) %>%
  mutate(label=ifelse(is.na(cmpd_name), formula, paste0(formula, " (", cmpd_name, ")"))) %>%
  mutate(casewhen_arg=paste0(cmd, "~'", label, "'"))
ms11_df <- c("C11H23NO6", "C11H23NO6", "C9H19NO5", "C8H17NO5", "C6H13NO6", "C8H17NO4") %>%
  data.frame(formula=., rtmed=c(8.5, 10, 8.5, 10.2, 5.2, 9.2)) %>%
  left_join(iter_mzs) %>%
  select(mzmed=mz, rtmed, formula, cmpd_name) %>%
  bind_rows(ms11_offenders) %>%
  arrange(mzmed) %>%
  select(mzmed, rtmed, formula, compound_name, feature) %>%
  mutate(formula=case_when(
    compound_name=="Taurine"~"C2H8NO3S",
    feature=="MS_POS_1050"~"C7H16NO2",
    compound_name=="Hydroxyisoleucine"~"C6H14NO3",
    compound_name=="N-Acetyltaurine"~"C4H10NO4S",
    feature=="MS_POS_1340"~"C5H14NO3S",
    feature=="MS_POS_1419"~"C8H18NO3",
    mzmed%between%pmppm(184.0637, 10)~"C5H14NO4S",
    mzmed%between%pmppm(198.0794, 10)~"C6H16NO4S",
    compound_name=="Dexpanthenol"~"C9H20NO4",
    TRUE~formula
  )) %>%
  mutate(compound_name=case_when(
    compound_name=="(3-Carboxypropyl)trimethylammonium?"~NA,
    compound_name=="Hydroxyisoleucine"~NA,
    compound_name=="Dexpanthenol"~NA,
    TRUE~compound_name
  ))
knitr::kable(ms11_df)
```

We explored this *Trichodesmium* hypothesis further by analyzing additional samples taken from the bloom to see whether the compounds abundant in *Trichodesmium* were also abundant in samples taken directly from the bloom itself. Of the 14 masses identified in the environmental sample, 7 had good-quality peaks when viewed in the bloom samples as extracted ion chromatograms, though only 4 of these matched closely in retention time.

```{r trichobloomsamples comparison, eval=FALSE}
library(RaMS)
trichodata <- list.files("../mzMLs/tricho_pos", pattern=".mzML", full.names = TRUE) %>%
  grabMSdata()
tricho_eics <- ms11_df %>%
  pmap(function(...){
  row_data <- list(...)
  peak_data <- trichodata$MS1[mz%between%pmppm(row_data$mzmed, 10)]
  cbind(peak_data, formula=row_data$formula)
}) %>%
  bind_rows()

tricho_eics %>%
  filter(rt%between%c(2, 15)) %>%
  arrange(mz) %>%
  mutate(formula=fct_inorder(formula)) %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename)) +
  facet_wrap(~formula, scales="free", ncol=2) +
  geom_vline(aes(xintercept = rtmed), color="red", data=ms11_df)

trichodata$MS1[mz%between%pmppm(162.1125, 10)] %>%
  mutate(samp_type=str_extract(filename, "Smp|Poo|Mix1|Mix2|Mat")) %>%
  filter(samp_type!="Mat") %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename, color=samp_type)) +
  ggtitle("Homoserine betaine in Tricho bloom samples")
```

Finally, we performed a search of the untargeted data for homoserine betaine, a known *Trichodesmium* osmolyte [@Pade2016]. Although this compound was not in our standard library due to a lack of purchasable authentic standards, we identified three mass features in the 162.1125 *m/z* window with retention times of 7.1, 8.4, and 9.9 minutes. The feature at RT 9.9 matched our carnitine standard's retention time, leaving the two earlier peaks with putative identifications of threonine betaine at 7.1 and homoserine betaine at 8.4 given the elution order of threonine before homoserine in our authentic standards. This order is the reverse of that in @Pade2016, which showed homoserine betaine eluting prior to carnitine as expected given our use of a polar HILIC column and their use of a nonpolar Hypercarb column. They did not report an additional peak detected at this mass, leaving our putative identification of the threonine betaine as low-confidence. None of the three 162.1125 features showed the same degree of enrichment in MS11C215m_A as taurine or the other 13 compounds detected with that pattern. Although MS11C215m_A was the largest, second-largest, and third-largest peak for carnitine, threonine betaine, and homoserine betaine, respectively, it was never more than triple the average surface value or double the next-largest value.

```{r homoserinebetaine search, eval=FALSE}
library(arrow)
dset <- open_dataset("../tmzMLs/MT_all_file_arr_pqds/")
hmb_eic <- dset %>%
  filter(mz%between%pmppm(162.1125, 10)) %>%
  # filter(mz%between%pmppm(120.0655, 10)) %>%
  collect()
hmb_eic %>%
  filter(rt%between%c(6, 11)) %>%
  # filter(rt%between%c(9, 13)) %>%
  mutate(samp_type=str_extract(filename, "Poo|Smp|InH2O|Mat")) %>%
  # distinct(filename, samp_type) %>%
  # filter(samp_type!="Smp")
  mutate(plotcol=ifelse(str_detect(filename, "180821_Smp_MS11C215m_A.mzML"), 
                          "MS11C215m_A", samp_type)) %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename, color=plotcol)) +
  facet_wrap(~samp_type, ncol=1, scales="free_y")

untarg_feats %>% 
  filter(cruise=="MS") %>%
  filter(mzmed%between%pmppm(162.1125))
untarg_peaks %>%
  filter(feature%in%c("MS_POS_1246", "MS_POS_1247", "MS_POS_1248")) %>%
  mutate(feature=factor(feature, levels=c("MS_POS_1248", "MS_POS_1246", "MS_POS_1247"),
                        labels=c("Threonine betaine (RT=7.1)", 
                                 "Homoserine betaine (RT=8.4)",
                                 "Carnitine (RT=9.9)"))) %>%
  left_join(filled_file_metadata) %>%
  filter(ms_run=="MT") %>%
  mutate(station=factor(station, levels=as.character(4:14))) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(plotcol=ifelse(filename=="180821_Smp_MS11C215m_A.mzML", "red", "black")) %>%
  ggplot(aes(x=station, y=norm_area, label=filename, color=plotcol)) +
  geom_point() +
  facet_grid(feature~depth, scales="free_y") +
  geom_hline(yintercept = 0) +
  scale_color_identity()

untarg_peaks %>%
  filter(feature%in%c("MS_POS_1246", "MS_POS_1247", "MS_POS_1248")) %>%
  left_join(filled_file_metadata) %>%
  filter(ms_run=="MT") %>%
  group_by(feature) %>%
  mutate(norm_area=norm_area/norm_area[filename=="180821_Smp_MS11C215m_A.mzML"]) %>%
  group_by(feature, depth) %>%
  summarise(mean_area=max(norm_area)) %>%
  pivot_wider(names_from = "depth", values_from = mean_area)
```

@@

```{r regression relative abundance coefficients, eval=FALSE}
reg_coefs <- untarg_peaks %>%
  # filter(feature!="MS_NEG_0155") %>%
  left_join(filled_file_metadata %>% distinct(filename, ms_run, station, depth, sla_corr)) %>%
  filter(ms_run=="MT") %>%
  filter(depth%in%c("Surface", "DCM", "Deep")) %>%
  filter(feature%in%unique(feature)) %>%
  group_by(filename) %>%
  mutate(norm_area=norm_area/sum(norm_area)) %>%
  ungroup() %>%
  nest(data=-c(depth, feature)) %>%
  mutate(test=map(data, ~broom::tidy(lm(.x$norm_area~.x$sla_corr)), .progress = TRUE)) %>%
  unnest(test)
reg_coefs %>%
  filter(str_detect(term, "sla_corr")) %>%
  mutate(p.value=p.adjust(p.value, method="fdr")) %>%
  arrange(p.value) %>%
  left_join(untarg_feats) %>%
  select(feature, depth, estimate, p.value, mzmed, rtmed, compound_name) %>%
  filter(p.value<0.05) %>%
  # as.data.frame() %>%
  filter(depth=="Deep") %>% 
  print(n=Inf)
reg_coefs %>%
  filter(str_detect(term, "sla_corr")) %>%
  left_join(untarg_feats) %>%
  filter(compound_name=="Homarine") %>%
  # filter(feature=="MS_POS_1395") %>%
  filter(depth=="DCM") %>%
  select(compound_name, depth, data, estimate) %>%
  unnest(data) %>%
  group_by(sla_corr, station) %>%
  ggplot(aes(x=sla_corr, y=norm_area*100)) +
  geom_smooth(method="lm", formula="y~x") +
  geom_boxplot(aes(group=sla_corr)) +
  scale_y_continuous(limits = c(0, NA))
```

@@

Finally, in the 175 meter samples we detected 43 mass features with a significant association with SLA. Notably, all detected nucleobases (guanine, adenine, and cytosine) were positively associated with SLA (higher values in the anticyclone than the cyclone), though the strongest associations were found in O-acetylcarnitine, betonicine, and tyrosine. Both O-acetylcarnitine and betonicine effectively quintupled their contribution to total peak area, shifting from approximately 0.1% to ~0.5% over the transect. Acetylcholine was the only known compound more relatively abundant in the cyclone than the anticyclone along with one unknown of *m/z* 131.034. The unknown at *m/z* 173.09211 with the strongest trend at the DCM again had the strongest trend in the 175 meter samples, here with an FDR-corrected p-value of $2.0*10^{-7}$.

@@

To provide context for the sea level anomaly responses detected above, we additionally performed high-resolution depth sampling around the DCM at the center of both eddy poles. Here, we found that the ~40 meter difference in depth and the ~40 centimeter difference in SLA explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001). Additionally, both sample depth and SLA were well-correlated with the first principal component of the metabolite matrix (r$_{depth}=.766$, r$_{SLA}=.684$, fraction of variance explained by PC$_1$ = 30.8%) (Figure 3). A 2D NMDS plot of the samples showed good separation between the eddies, with cyclonic samples largely on the left side of Figure 3 while anticyclonic samples fell almost entirely to the right. Notably, the cyclonic samples show a much larger spread than the anticyclonic ones, indicating larger sample variability at the surface than the DCM. A depth gradient is visible along MDS 2, with the deepest samples generally at the top right of the plot and the shallowest ones closer to the bottom and left.

```{r figure MC NMDS plot}
MC_nmds_gp <- multivar_analysis_MC$nmds_output %>%
  pluck(1, "points") %>% 
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_class, depth, tripl)) %>%
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"),
                      labels=c("-20m", "-10m", "DCM", "+10m", "+20m"))) %>%
  ggplot(aes(label=tripl)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, color=sla_class), size=4, pch=19) +
  scale_color_manual(
    values=c(scales::muted("blue"), scales::muted("red")),
    breaks=c("Cyclone", "Anticyclone"),
  ) +
  annotate("label", x=-200, y=100, label="Cyclone", color=scales::muted("blue"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=100, y=120, label="Anticyclone", color=scales::muted("red"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=-Inf, y=Inf, label=multivar_analysis_MC$nmds_stress,
           hjust=0, vjust=1, label.r = unit(0, "in")) +
  geom_point(aes(x=MDS1, y=MDS2, fill=depth, alpha=depth), size=4, pch=21, color="black") +
  scale_fill_manual(
    values = c("grey90", "grey80", "grey60", "grey40", "grey20"),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.2, 0, 0.2, 0.4),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  coord_fixed() +
  guides(alpha="none", color="none", fill=guide_legend(
    title = "Depth", title.position = "top", title.hjust = 0.5, direction = "horizontal"
  )) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave("MC_nmds_gp.png", plot = MC_nmds_gp, device = "png", 
       path = "figures", width = 6.5, height = 4.5, units = "in", dpi = 300)
```

@@

  - Discussion: I still really like the idea of modeling things according to the slope of their change across the eddy dipole. Light-sensitive or pressure-sensitive (e.g. sunscreen molecules, piezolytes) metabolites wouldn't change at all, DCM-specific metabolites (e.g. chlorophyll) would change by 20m or so as the DCM does, nutrient-sensitive metabolites (guanine? glutamate/glutamine?) would track with the isonutrials, and the salinity or temperature sensitive metabolites (aka pure osmolytes) would track with isopycnals. Do I have enough data to show that here? I forget what my early attempts to quantify these gradients looked like.

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 4). We found that a majority of the metabolites (Clusters 1 and 2, 63% of the total) fell into clusters with larger peak areas in the cyclone, as expected, while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 4). Cluster 2 notably had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster 1 also showed a neat DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (r=.851) that was not present in the cyclone (r=.013).

```{r k-means clusterplot}
set.seed(123)
kclust <- untarg_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MC") %>%
  select(feature, filename, norm_area) %>%
  group_by(feature) %>%
  filter(sum(norm_area==0)<n()*0.9) %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  pivot_wider(names_from = "filename", values_from = norm_area) %>%
  column_to_rownames("feature") %>%
  data.matrix() %>%
  kmeans(centers = 4)

@@

```{r eddy center volcano plot}
volcanoplot_gp <- pairwise_tests %>%
  left_join(fold_change_df, by = join_by(feature)) %>%
  left_join(rownames_to_column(data.frame(cluster=kclust$cluster), "feature"), 
            by = join_by(feature)) %>%
  left_join(untarg_feats %>% select(feature, compound_name), by = join_by(feature)) %>%
  mutate(known=ifelse(is.na(compound_name), "Unknown", "Known")) %>%
  mutate(compound_name=ifelse(is.na(compound_name), feature, compound_name)) %>%
  ggplot(aes(x=log2(fold_change), y=-log10(p.value), color=factor(cluster), 
             shape=known, label=compound_name)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 1.3, lty=2) +
  geom_point(size=2, stroke=1) +
  scale_y_continuous(breaks = 0:6, labels = paste0("10<sup>-", 0:6, "</sup>")) +
  scale_shape_manual(values=c(4, 16), breaks = c("Unknown", "Known")) +
  scale_color_manual(
    values = c("#414487FF", "#7AD151FF", "#22A884FF", "#2A788EFF"),
    breaks = 1:4
  ) +
  theme_bw() +
  labs(x="Log-2 fold-change cyclone enrichment", y="Mann-Whitney p-value\n(FDR-corrected)",
       shape="Metabolite\nstatus", color="k-means\ncluster") +
  guides(color=guide_legend(order = 2), shape=guide_legend(order=1)) +
  theme(axis.text.y = ggtext::element_markdown(),
        plot.background = element_rect(fill = "white"))
```

```{r kclust volcano plot combine}
kclust_volcano_gp <- egg::ggarrange(
  kclust_gp, volcanoplot_gp + theme(legend.position = "none"), 
  ncol=1, heights = c(2, 1.5)) %>%
  cowplot::plot_grid(
    cowplot::get_legend(volcanoplot_gp), rel_widths = c(0.85, 0.15)
  )
ggsave("kclust_volcano_gp.png", plot = kclust_volcano_gp, device = "png", 
       path = "figures", width = 6.5, height = 4, units = "in", dpi = 300,
       bg="white")
```

![](figures/kclust_volcano_gp.png)
